---
title: "Ch_11_Exercise"
output: html_document
---
Conduct the analyses for spatial dependence (11.17) using different cell types than those used in the book's code.
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library("readr")
library("dplyr")
library(ggplot2)
cellclasses = c("T_cells", "Tumor", "DCs", "other_cells")
brcalymphnode = lapply(cellclasses, function(k) {
    read_csv(file.path("..", "data",
             sprintf("99_4525D-%s.txt", k))) %>%
    transmute(x = globalX,
              y = globalY,
              class = k)
}) %>% bind_rows %>% mutate(class = factor(class))

brcalymphnode

ggplot(filter(brcalymphnode, class %in% c("T_cells", "Tumor")),
   aes(x = x, y = y, col = class)) + geom_point(shape = ".") +
   facet_grid( . ~ class) + guides(col = FALSE)
```

```{r}
library("spatstat")

ln = with(brcalymphnode,
  ppp(x = x, y = y, marks = class, xrange = range(x), yrange = range(y)))
ln

library("geometry")
coords = cbind(ln$x, ln$y)
chull = convhulln( coords )

pidx = integer(nrow(chull) + 1)
pidx[1:2] = chull[1, ]
chull[1, ] = NA
for(j in 3:length(pidx)) {
  wh = which(chull == pidx[j-1], arr.ind = TRUE)
  stopifnot(nrow(wh )== 1)
  wh[, "col"] = 3 - wh[, "col"] ## 2->1, 1->2
  pidx[j] = chull[wh]
  chull[wh[, "row"], ] = NA
}
pidx = rev(pidx)

ggplot(tibble(x = ln$x, y = ln$y)[pidx, ], aes(x = x, y = y)) +
  geom_point() + geom_path() + coord_fixed()

ln = with(brcalymphnode,
   ppp(x = x, y = y, marks = class, poly = coords[ pidx, ],
       check = FALSE))
ln

d = density(subset(ln, marks == "Tumor"), edge=TRUE, diggle=TRUE)
plot(d)

rr = relrisk(ln, sigma = 250)

plot(rr)

gln = Gest(ln)
gln
```

**It really starts here**

**Tumor**
```{r}
Lln = Linhom(subset(ln, marks == "Tumor"))
Lln

plot(Lln, lty = 1)

pcfln = pcf(Kinhom(subset(ln, marks=="Tumor")))

plot(pcfln, lty = 1)
plot(pcfln, lty = 1, xlim = c(0,10))
```

Therefore, the cells really close to each other repel a small amount (basically they aren't on top of each other), but they do cluster.

**DCs**

iso is where the intensity is the same regardless of the direction you're moving in
```{r}
Lln = Linhom(subset(ln, marks == "DCs"))
Lln

plot(Lln, lty = 1)

pcfln = pcf(Kinhom(subset(ln, marks=="DCs")))

plot(pcfln, lty = 1)
plot(pcfln, lty = 1, xlim = c(0,10))
```

These cells definitely cluster as g(r)>1.

**Other cells**
```{r}
Lln = Linhom(subset(ln, marks == "other_cells"))
Lln

plot(Lln, lty = 1)

pcfln = pcf(Kinhom(subset(ln, marks=="other_cells")))

plot(pcfln, lty = 1)
plot(pcfln, lty = 1, xlim = c(0,10))
```