---
title: "Chapter 12 Examples"
author: "Brooke Anderson"
date: "5/7/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r message = FALSE, warning = FALSE}
library(tidyverse)
```

## Diabetes example

```{r}
diabetes <- read_csv("../data/diabetes.csv")

diabetes %>% 
  slice(1:3)

diabetes %>% 
  dim()
```

This dataset has observations on 144 people, all adults who were not obese.
They've grouped these people into three groups (`group`): those who are "normal"
(non-diabetic), those with chemical diabetes, and those with overt diabetes.
Based on the paper they reference, for these subjects:

> "The variables used were age, relative weight, fasting plasma glucose, area
under the plasma glucose curve for the three hour oral glucose tolerance test,
area under the plasma insulin curve for the OGTT, and the steady state plasma
glucose response."

Based on this, I ny guesses for the columns in the dataframe are: 

- `id`: Patient unique identified
- `relwt`: Relative weight
- `glufast`: Fasting plasma glucose
- `glutest`: Area under the plasma gluocse curve for the three hour oral 
glucose tolerance test
- `steady`: The steady-state plasma glucose response
- `insulin`: Area under the plasma insulin curve for the OGTT
- `group`: Which of the three groups the subject belonged to (normal, 
chemical diabetes, or overt diabetes)

One question here is how each of the measured variables is associated with
the person's group status. Here's some code that does an adaptation of Figure 12.3
in the book:

```{r fig.height = 5, fig.width = 3, fig.align = "center"}
diabetes <- diabetes %>% 
  # Convert `group` to a factor and change the level names to be clearer
  # (what they actually represent)
  mutate(group = as_factor(group), 
         group = fct_recode(group, 
                            "Normal" = "1", 
                            "Chemical diabetes" = "2", 
                            "Overt diabetes" = "3"))
# Find out how many subjects are in each group
diabetes %>% 
  group_by(group) %>% 
  count()

diabetes %>% 
  # Reshape so we can plot everything with faceting
  pivot_longer(relwt:insulin) %>% 
  # Convert `name` (the new column created when you reshaped the data)
  # to a factor and replace with what they really are (for better labels in 
  # the graph)
  mutate(name = as_factor(name), 
         name = fct_recode(name, 
                           "Relative weight" = "relwt", 
                           "Fasting plasma glucose" = "glufast", 
                           "Plasma glucose under tolerance test" = "glutest", 
                           "Steady-state glusose response" = "steady", 
                           "Plasma insulin under tolerance test" = "insulin")) %>% 
  ggplot(aes(x = value, color = group)) + 
  geom_density() + 
  facet_wrap(~ name, scales = "free", ncol = 1) + 
  theme(legend.position = "bottom")
```

For this example dataset, I think we'll be trying to see if we can build a model
to predict a person's group status (normal, chemical diabetes, or overt diabetes) 
based on these measurements. 

They start by trying to predict `group` based on two measurements, `insulin` and `glutest`. 
Here's a scatterplot of those variables, with group status shown by color: 

```{r message = FALSE, warning = FALSE, fig.height = 3.5, fig.width = 5.5, fig.align = "center"}
# Calculate the means of each variable by group
# We'll add these as "X"s to show the group mean in the plot
group_means <- diabetes %>% 
  group_by(group) %>% 
  summarize_all(mean) 
group_means

# Since we're plotting from two dataframes, wait and add the "data" 
# when you add each geom.
ggplot() + 
  geom_point(data = diabetes, 
             aes(x = insulin, y = glutest, color = group)) + 
  geom_point(data = group_means, 
             aes(x = insulin, y = glutest), 
             shape = 4, size = 5) + 
  labs(x = "Plasma insulin under tolerance test", 
       y = "Plasma glucose under tolerance test", 
       color = "")
```

Try using LDA to build a predictive model based on these two measurements in the
data. Fit the model using `lda`:

```{r message = FALSE, warning = FALSE}
library(MASS)
diabetes_lda <- lda(group ~ insulin +  glutest, data = diabetes)
```

This has a special class, `lda`. Often, the output from fitting models in R have
special S4 (or S3) classes, which are essentially fancy lists.

```{r}
diabetes_lda %>% 
  class()
```

These will often have interesting `print` methods, which will show
some results from fitting the model: 

```{r}
# The `print` method runs by default when you run just the object's name
diabetes_lda
```

As with any object in the "list" family, you might be able to find out what's in the
object using `names` and `str`:

```{r}
diabetes_lda %>% 
  names()

diabetes_lda %>% 
  str()
```

You can use this model object to predict new class levels. You can do this by 
either for the original data or with new data that includes columns with `insulin` and
`glutest`. For example, here's a case of using this model to predict the subjects'
groups from the original data (of course, we know what the real values are!) based
on each subject's measures of `insulin` and `glutest` and on this model we just built.
By default, it predicts on the data used to fit the model, so we don't need to input that:

```{r}
diabetes_lda %>% 
  predict() %>% 
  # The` predict` object includes several elements. Right now, we just want to get the
  # classes
  pluck("class")
```

We will often want to compare that to the *true* classes for each study subject. 
Here's one way of doing that: 

```{r}
diabetes_lda %>% 
  predict() %>% 
  pluck("class") %>% 
  as_tibble() %>% 
  rename(predicted_group = value) %>% 
  # Add on a column with the real group labels (by default, the predictions will
  # be in the same order as the original data, so things should line up fine here)
  mutate(true_group = diabetes$group) %>% 
  # Get the numbers in each combo of true and predicted group labels
  group_by(predicted_group, true_group) %>% 
  count() %>% 
  # Make a plot to compare the true and predicted groups
  ggplot(aes(x = predicted_group, y = true_group, fill = n, label = n)) + 
  geom_tile() + 
  geom_text(color = "white") +
  labs(x = "Predicted group\n(based on LDA model)", 
       y = "Real group", 
       fill = "# of study\nsubjects") + 
  theme_classic()
```
 
To help visualize the predictions, they suggest that we make a regularly-spaced grid
all across the range of the original data (so it will fill up the background of 
the original scatterplot). Then, we'll use the LDA model to predict the class at each
of these grid points. This will let us see what the model would predict at each 
area of the original range of data. 

There's a wonderful function called `expand_grid` that lets you create a dataframe
with every possible combination of two or more sets of vectors. We can use that
to create a grid across the original plot area (we can use `min` and `max` on each variable
to figure out what that is).
 
```{r}
# Create a new dataframe with points to predict at. They'll be regularly-spaced
# points throughout the space of the original data
pred_grid <- expand_grid(insulin = seq(from = min(diabetes$insulin), 
                          to = max(diabetes$insulin), 
                          length.out = 100), 
            glutest = seq(from = min(diabetes$glutest), 
                          to = max(diabetes$glutest), 
                          length.out = 100))

pred_grid %>% 
  slice(1:5)

# This is just to show you that you do indeed get a regularly-spaced grid throughout
# the original graph area
ggplot(pred_grid, aes(x = insulin, y = glutest)) + 
  geom_point(size = 0.1)
```
 
Now we want to add on our predictions for each of these grid points, based on the LDA
model we built: 

```{r}
diabetes_lda %>% 
  # Predict the groups for each point in the expanded grid
  predict(newdata = pred_grid) %>% 
  pluck("class") %>% 
  as_tibble() %>% 
  # Add back in the values of `insulin` and `glutest` so we can plot
  # everything together
  mutate(insulin = pred_grid$insulin, 
         glutest = pred_grid$glutest) %>% 
  # Plot everything
  ggplot(aes(x = insulin, y = glutest, color = value)) + 
  geom_point(size = 0.1)
```

Now you can add back in the original data:

```{r}
diabetes_lda %>% 
  # Predict the groups for each point in the expanded grid
  predict(newdata = pred_grid) %>% 
  pluck("class") %>% 
  as_tibble() %>% 
  # Add back in the values of `insulin` and `glutest` so we can plot
  # everything together
  mutate(insulin = pred_grid$insulin, 
         glutest = pred_grid$glutest) %>% 
  # Plot the predictions for all the regions of the plot, based on the LDA model
  ggplot(aes(x = insulin, y = glutest, color = value)) + 
  geom_point(size = 1, alpha = 0.2) + 
  # Add in the observed data points
  geom_point(data = diabetes, 
             aes(x = insulin, y = glutest, color = group)) + 
  # Plot the points showing mean values of insulin and glutest for each group
  geom_point(data = group_means, 
             aes(x = insulin, y = glutest, color = NULL), 
             shape = 3, size = 5)
```

They discuss how this model is more likely, for observations between the group
means for `insulin` and `glutest` for groups 2 and 3, to predict 3 (overt diabetes)
rather than 2 (chemical diabetes). This is because the original data has a much 
higher proportion of observations in group 3 than group 2, which leads to the 
default prior weights for the LDA being skewed toward group 2. 

```{r}
# Calculate the total number of observations in each group, as well as
# the proportion of all observations in each group
diabetes %>% 
  group_by(group) %>% 
  count() %>% 
  ungroup() %>% 
  mutate(total = sum(n), 
         prop = n / total)

# Compare to the priors in the LDA model---you'll see they're just using 
# the proportion of observations in the group in the data used to train the
# model
diabetes_lda %>% 
  pluck("prior")
```

You can change re-fit an LDA model while forcing the prior weights for each 
group to be equal by adding weights for each observation. In essence, the aim is to 
"downweight" the class where we had more observations (overt obesity) and "upweight"
the other so that the priors are even between the three groups:

```{r}
diabetes_up <- lda(group ~ insulin + glutest, data = diabetes, 
                   prior = c(1/3, 1/3, 1/3))
diabetes_up
```

Create a plot with the predicted regions based on this new model, with a uniform prior:

```{r}
diabetes_up %>% 
  # Predict the groups for each point in the expanded grid
  predict(newdata = pred_grid) %>% 
  pluck("class") %>% 
  as_tibble() %>% 
  # Add back in the values of `insulin` and `glutest` so we can plot
  # everything together
  mutate(insulin = pred_grid$insulin, 
         glutest = pred_grid$glutest) %>% 
  # Plot the predictions for all the regions of the plot, based on the LDA model
  ggplot(aes(x = insulin, y = glutest, color = value)) + 
  geom_point(size = 1, alpha = 0.2) + 
  # Add in the observed data points
  geom_point(data = diabetes, 
             aes(x = insulin, y = glutest, color = group)) + 
  # Plot the points showing mean values of insulin and glutest for each group
  geom_point(data = group_means, 
             aes(x = insulin, y = glutest, color = NULL), 
             shape = 3, size = 5)
```

You might want to compare how well several models perform. One measure might be
the accuracy of each model in classifying the data---what percent of the observations
does the model assign the correct class? Here, we'll check that in the original 
data used to fit the model, but just note that that method of checking models can 
be prone to reward models that overfit to the training data. 

```{r}
# Fit your three models and put the results of each in the three slots in a list
list(two_var_unweighted = lda(group ~ insulin + glutest,
                                            data = diabetes),
     two_var_weighted = lda(group ~ insulin + glutest, 
                            data = diabetes,
                            prior = c(1/3, 1/3, 1/3)), 
     five_var_unweighted = lda(group ~ relwt + glufast + glutest + 
                                 steady + insulin, data = diabetes)
     ) %>% 
  # Use `map` to run `predict` for each of the model objects
  map(predict) %>% 
  # Pluck out the first element of the output of predict for each (the class predictions)
  map(1) %>% 
  # Convert this from a vector to a one-column dataframe
  map(as_tibble) %>% 
  # Add in a column with the true values of each observation (from the original data)
  map(~ mutate(., true_class = diabetes$group)) %>% 
  # Stick everything together into one big data frame
  bind_rows(.id = "model") %>% 
  # Check for when the prediction is right (i.e., the same as the true group
  # based on the original data)
  mutate(right = value == true_class) %>% 
  # Get summaries of what percent of the time the prediction is right for each 
  # model (and what percent of the time it's wrong, since that's what they 
  # calculated in the book)
  group_by(model) %>% 
  summarize(perc_right = round(100 * mean(right), 1), 
            perc_wrong = round(100 * mean(!right), 1)) %>% 
  mutate(model = fct_recode(model, 
                            "Unweighted, all 5 variables" = "five_var_unweighted", 
                            "Unweighted, insulin and glucose test" = "two_var_unweighted", 
                            "Weighted, insulin and glucose test" = "two_var_weighted")) %>% 
  knitr::kable(col.names = c("Model", 
                             "Percent predicted correctly", 
                             "Percent predicted incorrectly"))
```

## Embryonic cell state example

They give another example, where they have some gene expression data from measurements
made at several different embryonic days, and they want to see if they can build a 
model that uses expression levels for four of these genes to predict the development
stage (embryonic day). 

Load the data:

```{r}
library("Hiiragi2013")
data("x")

x %>% 
  class()
```

This data is in an `ExpressionSet` class. You can extract the expression levels
with the `exprs` accessor method: 

```{r}
embryonicDay <- x %>% 
  # Extract expression levels
  exprs() %>% 
  # Transpose, so that samples are on rows and gene expression levels
  # in columns
  t() %>% 
  # Put the rownames in a column, because we want to keep that info (it
  # helps identify each sample). You can do this while you convert to 
  # a tibble.
  as_tibble(rownames = "sample_id") %>% 
  # Keep just the measurements for the four genes that you care about
  dplyr::select(sample_id, `1426642_at`, `1418765_at`, `1418864_at`, `1416564_at`) %>% 
  # Use regular expressions to pull out the embryonic day from the sample IDs
  mutate(embry_day = str_extract(sample_id, "E.+"),
         embry_day = str_remove(embry_day, "\\(.+"),
         embry_day = str_squish(embry_day),
         embry_day = as_factor(embry_day))

embryonicDay %>% 
  dplyr::slice(1:3)
  
```

At this point, we have a much smaller dataframe, with data for each sample giving
the sample ID, the expression levels of four genes, and the embryonic day. We're going
to try to build a model that predicts embryonic day (`embry_day`) based on the four
gene expression levels.

However, it sounds like we also want to use the probe IDs for these genes to get
symbols and gene names for each of them. To do that, it sounds like we can first use
the `annotation` function to figure out what annotation package to use for this data:

```{r}
x %>% 
  annotation()
```

Once we know that, we can load the right package and then get the annotation dataset:

```{r message = FALSE, warning = FALSE}
library("mouse4302.db")
probe_annos <- mouse4302.db %>% 
  AnnotationDbi::select(keys = embryonicDay %>% colnames() %>% str_subset("^[0-9]"),
                        columns = c("SYMBOL", "GENENAME"))

probe_annos
```

Now we can merge this in with out original data to get some better column names: 

```{r}
embryonicDay %<>%
  # Lengthen the data so we can join with the better symbols and replace these
  # as the column names, then we'll pivot back out to this shape
  pivot_longer(-c(sample_id, embry_day)) %>% 
  full_join(probe_annos %>% dplyr::select(PROBEID, SYMBOL), 
            by = c("name" = "PROBEID")) %>% 
  dplyr::select(-name) %>% 
  pivot_wider(names_from = SYMBOL, values_from = value)

# Yah, better column names!
embryonicDay %>% 
  dplyr::slice(1:3)
```

Investigate associations between these four gene expression levels and the 
embryonic day:

```{r message = FALSE, warning = FALSE}
library(GGally)

embryonicDay %>% 
  ggpairs(aes(color = embry_day),
          # Only show for the four gene expression columns
          columns = 3:6,
          upper = list(continuous = "points"))

```

Fit an LDA predictive model: 

```{r}
ec_lda <- lda(embry_day ~ Fn1 + Timd2 + Gata4 + Sox7, 
              data = embryonicDay)
ec_lda
```

Check the linear combinations LD1 and LD2 that serve as discriminating variables:

```{r}
ec_lda %>% 
  pluck("scaling") %>% 
  round(1)
```


## Data for the exercise

The `ElemStatPackage` has been orphaned and isn't on CRAN anymore. However, it's up on
GitHub, so I grabbed the data file you'll need from there. You can download it 
yourself at: https://github.com/cran/ElemStatLearn/blob/master/data/prostate.RData

```{r}
load("../data/prostate.RData")

prostate %>% 
  head()
```

Here's a description of the data, from the archived help files: 

> "Data to examine the correlation between the level of prostate-specific
  antigen and a number of clinical measures in men who were about to 
  receive a radical prostatectomy."

Here's what the variables mean: 

- `lcavol`: log cancer volume
- `lweight`: log prostate weight
- `age`: in years
- `lbph`: log of the amount of benign prostatic hyperplasia
- `svi`: seminal vesicle invasion
- `lcp`: log of capsular penetration
- `gleason`: a numeric vector with the Gleason score
- `pgg45`: percent of Gleason score 4 or 5
- `lpsa`: response (the thing you are trying to predict), the 
level of prostate-specific antigen
- `train`: a logical vector, of whether the data was to be 
part of the training dataset (TRUE) or the testing one (FALSE)

So, you're trying to predict the values of `lpsa` based on the variables
`lcavol` through `pgg45`. 

In this exercise, they ask you to use the `glmnet` package:

```{r}
# install.packages("glmnet")
library(glmnet)
```

For `glmnet`, I think you need to input a matrix of the predictors (`x`) and, in the 
same order, a vector with the outcomes (`y`). So, here's an example of building a 
model to predict `lpsa` from `prostate` based on all the variables between `lcavol` and
`pgg45`:

```{r}
mod_basic <- glmnet(x = prostate %>% dplyr::select(lcavol:pgg45) %>% as.matrix(), 
                    y = prostate %>% pull(lpsa), 
                    family = "gaussian")

mod_basic 
```

