---
title: "Ch_5"
author: "Amy Fox"
date: "3/4/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Binary distance
Looking at strain 1 & 2
Find the number of total genes that are "on" (so this would be there are 5 genes that are 1's) - there are 5 total different genes that are on in at least 1 strain

There is 1 that is on for both of the strains, so we get the value as 1/5 --> to calculate the distance, we take 1-1/5 = 4/5 = 0.8

A larger number means that their further apart in distance

# 5.5.2 Data Preprocessing

You can visualize flow data with the flowViz package
```{r}
library("flowCore") 
library("flowViz") 
fcsB = read.FCS("../data/Bendall_2011.fcs") 
slotNames(fcsB)

markersB = readr::read_csv("../data/Bendall_2011_markers.csv")
mt = match(markersB$isotope, colnames(fcsB))
stopifnot(!any(is.na(mt)))
colnames(fcsB)[mt] = markersB$marker


flowPlot(fcsB, plotParameters = colnames(fcsB)[2:3], logy = TRUE)
```

It's necessary to transform your flow data in order to see clusters of cells. Arcsinh is used because it can deal with negative numbers.
```{r}

asinhtrsf = arcsinhTransform(a = 0.1, b = 1)
fcsBT = transform(fcsB,
  transformList(colnames(fcsB)[-c(1, 2, 41)], asinhtrsf))
densityplot( ~`CD3all`, fcsB)
densityplot( ~`CD3all`, fcsBT)
```

Combine contour plot with the flow plot
```{r}
kf = kmeansFilter("CD3all" = c("Pop1","Pop2"), filterId="myKmFilter")
fres = flowCore::filter(fcsBT, kf)
summary(fres)

fcsBT1 = flowCore::split(fcsBT, fres, population = "Pop1")
fcsBT2 = flowCore::split(fcsBT, fres, population = "Pop2")

flowPlot(fcsBT, plotParameters = c("CD3all", "CD56"), logy = FALSE)
contour(fcsBT[, c(40, 19)], add = TRUE)
```

```{r}
library("ggcyto")
library("labeling")
ggcd4cd8=ggcyto(fcsB,aes(x=CD4,y=CD8))
ggcd4=ggcyto(fcsB,aes(x=CD4))
ggcd8=ggcyto(fcsB,aes(x=CD8))
p1=ggcd4+geom_histogram(bins=60)
p1b=ggcd8+geom_histogram(bins=60)
asinhT = arcsinhTransform(a=0,b=1)
transl = transformList(colnames(fcsB)[-c(1,2,41)], asinhT)
fcsBT = transform(fcsB, transl)
p1t=ggcyto(fcsBT,aes(x=CD4))+geom_histogram(bins=90)
p2t=ggcyto(fcsBT,aes(x=CD4,y=CD8))+geom_density2d(colour="black")
p3t=ggcyto(fcsBT,aes(x=CD45RA,y=CD20))+geom_density2d(colour="black")

```

```{r}
library("dbscan")
mc5 = Biobase::exprs(fcsBT)[, c(15,16,19,40,33)]
res5 = dbscan::dbscan(mc5, eps = 0.65, minPts = 30)
mc5df = data.frame(mc5, cluster = as.factor(res5$cluster))
table(mc5df$cluster)

ggplot(mc5df, aes(x=CD4,    y=CD8,  col=cluster))+geom_density2d()
ggplot(mc5df, aes(x=CD3all, y=CD20, col=cluster))+geom_density2d()
```



# Exercise 5.1

A) 
```{r}
library("cluster")
library(dplyr)
library(ggplot2)

set.seed(23)

# copied from section 5.7
simdat = lapply(c(0, 8), function(mx) {
  lapply(c(0,8), function(my) {
    tibble(x = rnorm(100, mean = mx, sd = 2),
           y = rnorm(100, mean = my, sd = 2),
           class = paste(mx, my, sep = ":"))
   }) %>% bind_rows
}) %>% bind_rows


simdatxy = simdat[, c("x", "y")] # without class label


pam4 <- pam(simdatxy, 4)

sil <- silhouette(pam4, 4)

plot(sil, col = c("red", "green", "blue", "purple"), main = "Silhouette")

summary(sil)$avg.width
```

Silhoutte index = 0.52


B) 
```{r}
library(purrr)
k <- c(2:10)

df_test <- data.frame()

for (i in 2:10){
  
  pam_run <- pam(simdatxy, i)
  sil_run <- silhouette(pam_run, i)
  
  row_to_add <- data.frame(i, width = summary(sil_run)$avg.width)
  
  df_test <- rbind(df_test, row_to_add)
}

df_test

ggplot(df_test, aes(i, width)) +
  geom_point() +
  geom_line() +
  xlab("k") +
  ylab("Silhouette Index") +
  ggtitle("Testing different k values for Silhouette Index")


summary(sil_run)
```

Because the Silhouette index is the highest at k = 4, then this should be the best silhouette index.

C) 

```{r}

set.seed(23)
# copied from section 5.7 but then changed the groups to 1 instead of 4 groups
simdat_diff = lapply(c(1), function(mx) {
  lapply(c(1), function(my) {
    tibble(x = rnorm(100, mean = mx, sd = 2),
           y = rnorm(100, mean = my, sd = 2),
           class = paste(mx, my, sep = ":"))
   }) %>% bind_rows
}) %>% bind_rows


simdatxy_adjust = simdat_diff[, c("x", "y")] # without class label

ggplot(simdatxy_adjust, aes(x = x, y = y)) +
  geom_point()

pam4 <- pam(simdatxy_adjust, 4)

sil <- silhouette(pam4, 4)

plot(sil, col = c("red", "green", "blue", "purple"), main = "Silhouette")
```

