---
title: "Ch_7"
author: "Amy Fox"
date: "3/25/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ade4)
library(factoextra)
library(dplyr)
```

# 7.8
```{r}
load("../data/Msig3transp.RData")
pcaMsig3 = dudi.pca(Msig3transp, center = TRUE, scale = TRUE,
                    scannf = FALSE, nf = 4)
fviz_screeplot(pcaMsig3) + ggtitle("")

ids = rownames(Msig3transp)
celltypes = factor(substr(ids, 7, 9))
status = factor(substr(ids, 1, 3))
table(celltypes)

cbind(pcaMsig3$li, tibble(Cluster = celltypes, sample = ids)) %>%
ggplot(aes(x = Axis1, y = Axis2)) +
  geom_point(aes(color = Cluster), size = 5) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_vline(xintercept = 0, linetype = 2) +
  scale_color_discrete(name = "Cluster") + coord_fixed()
```

#7.8.1
```{r}
load("../data/mat1xcms.RData")
dim(mat1)

pcamat1 = dudi.pca(t(mat1), scannf = FALSE, nf = 3)
fviz_eig(pcamat1, geom = "bar", bar_width = 0.7) + ggtitle("")

dfmat1 = cbind(pcamat1$li, tibble(
    label = rownames(pcamat1$li),
    number = substr(label, 3, 4),
    type = factor(substr(label, 1, 2))))
pcsplot = ggplot(dfmat1,
  aes(x=Axis1, y=Axis2, label=label, group=number, colour=type)) +
 geom_text(size = 4, vjust = -0.5)+ geom_point(size = 3)+ylim(c(-18,19))
pcsplot + geom_hline(yintercept = 0, linetype = 2) +
  geom_vline(xintercept = 0, linetype = 2)
```

# 7.8.2
```{r}
library("pheatmap")
load("../data/wine.RData")
load("../data/wineClass.RData")
wine[1:2, 1:7]

pheatmap(1 - cor(wine), treeheight_row = 0.2)


winePCAd = dudi.pca(wine, scannf=FALSE)
table(wine.class)

fviz_pca_biplot(winePCAd, geom = "point", habillage = wine.class,
   col.var = "violet", addEllipses = TRUE, ellipse.level = 0.69) +
   ggtitle("") + coord_fixed()
```

# 7.8.3
```{r}
#BiocManager::install("Hiiragi2013")


data("x", package = "Hiiragi2013")
xwt = x[, x$genotype == "WT"]
sel = order(rowVars(Biobase::exprs(xwt)), decreasing = TRUE)[1:100]
xwt = xwt[sel, ]
tab = table(xwt$sampleGroup)
tab

xwt$weight = 1 / as.numeric(tab[xwt$sampleGroup])
pcaMouse = dudi.pca(as.data.frame(t(Biobase::exprs(xwt))),
  row.w = xwt$weight,
  center = TRUE, scale = TRUE, nf = 2, scannf = FALSE)
fviz_eig(pcaMouse) + ggtitle("")

fviz_pca_ind(pcaMouse, geom = "point", col.ind = xwt$sampleGroup) +
  ggtitle("") + coord_fixed()
```

# Exercise 7.4
```{r}
library(tidyverse)
library(Biobase)

data("x", package = "Hiiragi2013")

# Unweighted PCA
```

## Weighted PCA example (7.8.3)

```{r}
# Uncomment and run the next line if you don't have the `Hiiragi2013` package
# BiocManager::install("Hiiragi2013")

data("x", package = "Hiiragi2013")
class(x)
str(x, max.level = 2)
x@phenoData # Remember that you can use `@` to extract an element from an S4 object
str(x@phenoData)

x %>% 
  exprs() %>% 
  `[`(1:10, 1:10) # Using a trick to pipe into the `[ , ]`-style subsetting function
                  # It's an "in-fix" function (like "+" and "/"), so you usually use
                  # it within a line of code (instead of with parentheses). However, 
                  # all of those will also work like regular functions if you surround
                  # the name with backticks, so '1 + 2' is the same in R as '`[`(1, 2)'.
```


Functions like `exprs`, which exist only to extract some of the data stored in a certain
type of object, are called *extractor* functions. If you need to get all the way to a 
dataframe, run `as_tibble` (from the tidyverse) or `as.data.frame` (from base R) right 
after you extract these data. 

You can use `ggplot2` to explore the data, although keep in mind that the column names
aren't in a standard formula. Instead, they start with numbers and have spaces:

```{r}
x %>% 
  exprs() %>% 
  colnames() %>% 
  head()
```

That means that you'll need to "protect" the column name in any tidyverse code, 
by using backticks around the column name. For example, you can run the following to 
create a histogram of expression levels in the first column (the first animal sample?): 

```{r fig.align = "center", fig.width = 3, fig.height = 2}
x %>% 
  exprs() %>% 
  as_tibble() %>% 
  ggplot(aes(x = `1 E3.25`)) + 
  geom_histogram(bins = 100)
```

If you'd like to get histograms for several of these, you can take advantage of 
`pivot_longer` and facetting to do that. For example, to create these for the 
a random sample of twelve columns, run: 

```{r}
x %>% 
  exprs() %>% 
  as_tibble() %>% 
  select(sample(1:ncol(.), size = 12)) %>%  # Sample twelve columns. The `.` is a "pronoun"--
                                            # it refers to the dataframe you've just piped in
  pivot_longer(cols = 1:12, names_to = "sample", values_to = "level") %>% 
  ggplot(aes(x = level)) + 
  geom_histogram(bins = 100) + 
  facet_wrap(~ sample)
```

It looks like each of the original column names includes "FGF4-KO" if the sample animal is
a mutant, rather than a wild-type. We might want to use the fill of the bars to show which
samples are wild-type versus mutant. Once you've made the data longer, you can use
regular expressions to determine, based on whether a label includes "FGF4-KO", if the
animal is a mutant, and then use that when you plot:

```{r}
x %>% 
  exprs() %>% 
  as_tibble() %>% 
  select(sample(1:ncol(.), size = 12)) %>% 
  pivot_longer(cols = 1:12, names_to = "sample", values_to = "level") %>% 
  mutate(mutant = str_detect(sample, "(FGF4-KO)")) %>% # Use regular expressions here
  ggplot(aes(x = level, fill = mutant)) +  # Add the mapping to fill for the `mutant` column
  geom_histogram(bins = 100) + 
  facet_wrap(~ sample) + 
  theme(legend.position = "bottom")
```

It looks like each row is for a separate gene (? mRNA? transcript?). You might want to 
instead get histograms for each of those (instead of by sample). I think the easiest way
to do that would be to transpose the data first (`t`---that is, flip the rows and columns)
and then continue from there: 

```{r}
x %>% 
  exprs() %>% 
  t() %>% # Here's where I'm switching rows and columns
  as_tibble() %>% 
  select(sample(1:ncol(.), size = 12)) %>% 
  pivot_longer(cols = 1:12, names_to = "transcript", values_to = "level") %>% 
  ggplot(aes(x = level)) +  
  geom_histogram(bins = 100) + 
  facet_wrap(~ transcript) 
```

Correlation matrices might be interesting here, too. The `ggcorrplot` package has some 
nice functions for making those. First, check out the size of the data: 

```{r}
x %>% 
  exprs() %>% 
  dim()
```

We could probably fit 101 values in a correlation plot, so we could do one of all the 
samples, but we probably can't for all the gene expression levels (over 45,000!). We can
look to see correlation patterns in that, but we probably should look at just a sample, not
everything at once.

```{r}
library(ggcorrplot)
x %>% 
  exprs() %>% 
  t() %>% 
  as_tibble() %>% 
  select(sample(1:ncol(.), size = 100)) %>% 
  cor() %>%  # Calculate the correlation matrix
  ggcorrplot(outline.col = "white", type = "upper") + 
  theme_void() # No point in having x and y labels right now---they'll be too small to see
```

It looks like you probably have some columns that are pretty strongly correlated with 
each other, both negatively and positively.

## Tidier version of code in book

The code in the book walks you through doing a weighted PCA. They first recommend
that you limit the data to the wild-type samples and then select the 100 features (genes?)
with the highest overall variance. Here's a "tidier" way to do that than in the book.

First, you can use one pipeline to make a dataframe that's limited to the 66 wild-type
samples and the top 100 features by variance. We're transposing it along the way (`t`), 
so it will be in the right format for the `dudi.pca` call later:

```{r}
simpl_data <- x %>% 
  exprs() %>% 
  t() %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "sample") %>% 
  filter(!str_detect(sample, "(FGF4-KO)")) %>% 
  pivot_longer(-sample, names_to = "transcript", values_to = "level") %>% 
  group_by(transcript) %>% 
  nest() %>% 
  mutate(var = map_dbl(.x = data, .f = ~ var(.x$level))) %>% # Calculate the variance for 
                                                             # each transcript
  ungroup() %>% 
  top_n(n = 100, wt = var) %>% # Extract the top 100 transcripts in terms of variance 
  select(-var) %>% # We don't need the variance now that we've picked the top 100, go remove it
  unnest(data) %>% # Unnest the data
  pivot_wider(names_from = transcript, values_from = level) # Make wide again
simpl_data
```

You can see our new data frame has 66 samples (just the wild-type samples) and 101 columns
(the sample names, then the 100 genes with the largest variance in expression levels).

We can pull out the group from them (in each sample name, everything from "E" later)
using regular expressions:

```{r}
simpl_data <- simpl_data %>% 
  mutate(group = str_extract(sample, "E.+")) # Pull everything starting from "E" in 'sample'
simpl_data %>% 
  select(sample, group)
```

Next, they want us to create a weight for each sample, as the inverse of how many total 
samples there are in its group. 

```{r}
simpl_data <- simpl_data %>% 
  group_by(group) %>% # Group by 'group' so we can get the total count for each group
  mutate(n_in_group = n(), # When things are grouped, `mutate` will *add* the summary
                           # information as a column, while keeping the same number of rows
                           # as the original 
         weight = 1 / n_in_group) %>% # We want the weight to be 1 / the number of groups
  ungroup()
# Here's an example of what a few of these look like now:
simpl_data %>% 
  select(sample, group, n_in_group, weight) %>% 
  sample_n(6)
```

```{r}
library(ade4)
pcaMouse <- simpl_data %>% 
  select(-sample, -group, -n_in_group, -weight) %>% 
  dudi.pca(center = TRUE, scale = TRUE, nf = 2, scannf = FALSE, 
           row.w = simpl_data$weight)
pcaMouse
library(factoextra)
pcaMouse %>% 
  fviz_eig() + 
  ggtitle("")
```

```{r}
pcaMouse %>% 
  fviz_pca_ind(geom = "point", col.ind = simpl_data$group) + 
  ggtitle("") + 
  coord_fixed() 
```


E: Embryonic day
genes as columns

```{r}
colnames(simpl_data)

simpl_data %>%
  select(group, n_in_group) %>%
  unique()

for_7_8 <- simpl_data %>%
  select(-sample, -group, -n_in_group, -weight)

pca_unweight <- dudi.pca(for_7_8, center = TRUE, scale = TRUE, scannf = FALSE)

fviz_eig(pca_unweight, geom = "bar", bar_width = 0.7) +
  ggtitle("Unweighted")

fviz_pca_biplot(pca_unweight, geom = "point")

fviz_pca_var(pca_unweight)

```
Can try downsampling 

```{r}
# weighted data
weighted_data <- simpl_data %>%
  group_by(group) %>%
  sample_n(size = 4) %>%
  ungroup()

weighted_for_pca <- weighted_data %>%
  select(-sample, -group, -n_in_group, -weight)

colnames(weighted_for_pca)

pca_weight <- dudi.pca(weighted_for_pca, center = TRUE, scale = TRUE, scannf = FALSE)

fviz_eig(pca_weight, geom = "bar", bar_width = 0.7) +
  ggtitle("Weighted")

fviz_pca_biplot(pca_weight, geom = "point")

# could create a variable for 
fviz_pca_var(pca_weight)

# c1 first principal component 
pca_weight$c1 %>%
  rownames_to_column("genes") %>%
  mutate(new = sqrt((CS1^2) + (CS2^2))) %>%
  arrange(desc(new))

```
The most correlated by both the principal components: 1420086_x_at

```{r}

top_10 <- pca_weight

top_10$c1 <- pca_weight$c1 %>%
  arrange(desc(CS1)) %>%
  top_n(10)

fviz_pca_biplot(top_10, geom = "point")

```

