---
title: "Ch.1"
author: "Amy Fox"
date: "1/27/2020"
output: html_document
---

All notes are from the book Modern Statistics for Modern Biology

http://web.stanford.edu/class/bios221/book/Chap-Generative.html


Additional notes: always set the seed

If p is small, then Poisson is better 

The beauty of this simulation is that you can see the probability of seeing a false positive in your own research before you even take the data - you just need to know the sample size, false positive rate, and number of observations 

You could also bootstrap with your current data to calculate if your data could actually be due to random noise

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Want to see the proability that 3 mutations would occur in an HIV strain based on a Poisson(5) model
```{r}
dpois(x = 3, lambda = 5)
```

The probability of seeing 3 mutations is `r dpois(x = 3, lambda = 5)` 

Now if we want to see the probability of seeing 0:12 mutations we could do

```{r}
dpois(x = 0:12, lambda = 5)
barplot(dpois(0:12, 5), names.arg = 0:12, col = "red")
```
Based on the barplot we can see that we will often see 4-5 mutations


Dealing with factors:
```{r}
genotype = c("AA","AO","BB","AO","OO","AO","AA","BO","BO",
             "AO","BB","AO","BO","AB","OO","AB","BB","AO","AO")

table(genotype)

# convert to factors
genotypeF = factor(genotype)
levels(genotypeF)
table(genotypeF)
```

```{r}
pvec = c(1/8, 3/8, 3/8, 1/8)
rmultinom(4, prob = pvec, size = 1)
```


15 coin tosses with an equal coin (0.5 probability) where in size, we say that we're only going to toss 2 coins
```{r}
rbinom(15, prob = 0.5, size = 1)
```

If we want to say that the probability of a ball falling into a blue box is 2/3 and a red box 1/3
```{r}
rbinom(15, 2/3, size = 1)
```

In this case, a 1 indicates the the ball fell into the blue box.

If we want to see how many times the ball falls into the blue box we could do
```{r}
set.seed(202)
rbinom(1, prob = 2/3, size = 12)
```

10 balls would fall into the blue box out of 12 throws

Probability mass distribution
```{r}
probabilities = dbinom(0:15, prob = 0.3, size = 15)
round(probabilities, 2)
barplot(probabilities, names.arg = 0:15, col = "red")
```

Probability can be approximated with Poisson distribution with the number of trials in large and the success of p is small.

Test the probabilty of :
```{r}
rbinom(n = 1, prob = 5*10^-4, size = 10000)

sim <- rbinom(n = 3000, prob = 5*10^-4, size = 10000)
barplot(table(sim),  col = "light blue")
```

**Not sure if I did this task correctly**

Find the sum of 50 independent Bernoulli variables with p = 0.01 

Is this the same as poisson 0.5 random variable

Bernoulli = 1 shot (with probabilty p)
The summation of the bournouli's = binomial 

Do 1000 simulations to compare the distributions
```{r}
set.seed(100)
n = 50
p = 0.01
sims = 100000

x <- rbinom(n = sims, prob = p, size = n) # binomial 

# multiply Poisson by simulation so that the binomial is in the same y-axis
hist(x)
points(0:6, dpois(0:6, n*p)*sims, col = "red")
lines(0:6, dpois(0:6, n*p)*sims, col = "red")


# Therefore, I may not want to use this approximation
# TO fix this, we could make p bigger or n smaller 
```

What is the probability that 7 or more epitopes would be found when the Poisson probability is 0.5

P(X>=7) --> P(X<6)
```{r}
# can calculate this way 
1-ppois(6, 0.5)

# or this way
ppois(6, 0.5, lower.tail = FALSE)
```


Monte Carlo

100,000 instances of picking the maximum of 100 Poisson distributions
```{r}
maxes = replicate(100000, {
  max(rpois(100, 0.5))
})

table(maxes)
```

Exercise 1.8
```{r}

```

You can kind of make up statistical equations to simulate - for example, could just do summation of absoluate value of expected - observed.
```{r}
library(BSgenome.Celegans.UCSC.ce2)

library(Biostrings)

genome_length <- length(Celegans$chrM)

# probabilities
probs <- letterFrequency(Celegans$chrM, letters = c("A", "C", "G", "T"))/length(Celegans$chrM)

set.seed(2)
pvec <- rep(1/4, 4)
obs <- rmultinom(1000, prob = pvec, size = genome_length)

expected0 <- pvec*genome_length

    
stat = function(obsvd, exptd = genome_length * pvec) {
  sum((obsvd - exptd)^2 / exptd)
}
stat(obs[, 1])


S0 = apply(obs, 2, stat)
summary(S0)


hist(S0, breaks = 25, col = "lavender", main = "")

# this gives us the test statistic
q95 = quantile(S0, probs = 0.95)
q95

# if we set an alpha of 0.05, then 95% of the data should be 

# Because our stat = 4392 (below) is much greater than our quantile of 8.045, then we reject the null (the probability is not uniform)

# to get the 95% quantile where 3 is 1 fewer the number of bins you have
qchisq(0.95, 3)
```

Don't need this
```{r}

obs <- rmultinom(1000, prob = probs, size = genome_length)

expected0 <- pvec*genome_length

    
stat = function(obsvd, exptd = genome_length * pvec) {
  sum((obsvd - exptd)^2 / exptd)
}
# this gives us the observed values
stat(obs[, 1])


S0 = apply(obs, 2, stat)
summary(S0)

hist(S0, breaks = 25, col = "lavender", main = "")

q95 = quantile(S0, probs = 0.95)
q95


```

